---
title: "Assignment 3: ML exercise"
format: html
author: Jing Tao
---


## 1. Review ISLR Chapter 3

## 2. Run Lab_ML01.R

# ============================================================================
# Workshop: Machine Learning with Survey Data (TEDS 2016)
# EPPS 6323 Knowledge Mining
# Karl Ho, University of Texas at Dallas
# ============================================================================

# ----------------------------------------------------------------------------
# 0. Setup: Packages and Data
# ----------------------------------------------------------------------------
```{r}
library(haven)        # Read Stata files
library(tidyverse)    # Data wrangling and visualization
library(GGally)       # Pairs plots
library(cluster)      # Clustering
library(factoextra)   # Visualize clusters and PCA
library(rpart)        # Decision trees
library(rpart.plot) # Plot decision trees
library(randomForest) # Random forest
library(caret)        # Model training and evaluation
library(e1071)        # SVM and Naive Bayes

# Load data
TEDS_2016 <- read_stata("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")


# ============================================================================
# PART I: EXPLORATORY DATA ANALYSIS
# ============================================================================

# ----------------------------------------------------------------------------
# Q1. Data overview
# ----------------------------------------------------------------------------

dim(TEDS_2016)
names(TEDS_2016)
summary(TEDS_2016)

# ----------------------------------------------------------------------------
# Q2. Missing data assessment
# ----------------------------------------------------------------------------

TEDS_2016 %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

# Visualize missingness
TEDS_2016 %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  ggplot(aes(x = reorder(variable, n_missing), y = n_missing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Missing Values by Variable",
       x = NULL, y = "Count of Missing Values") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q3. Variable recoding
# ----------------------------------------------------------------------------

teds <- TEDS_2016 %>%
  mutate(
    vote = factor(votetsai, 
                  levels = c(0, 1), 
                  labels = c("Other", "Tsai")),
    gender = factor(female, 
                    levels = c(0, 1), 
                    labels = c("Male", "Female")),
    Tondu = as.factor(Tondu),
    Party = as.factor(Party)
  ) %>%
  dplyr::select(vote, gender, age, edu, income, 
         Taiwanese, Econ_worse, Tondu, Party, DPP) %>%
  drop_na()

glimpse(teds)

# ----------------------------------------------------------------------------
# Q4. Vote choice distribution
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = vote, fill = vote)) +
  geom_bar() +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Vote Choice Distribution",
       subtitle = "TEDS 2016: Tsai Ing-wen vs. Other Candidates",
       x = "Vote Choice", y = "Count") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")

# ----------------------------------------------------------------------------
# Q5. Age distribution
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Age Distribution of Respondents",
       x = "Age", y = "Count") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q6. Vote by gender
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = gender, fill = vote)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Vote Choice by Gender",
       x = "Gender", y = "Proportion",
       fill = "Vote") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q7. Age by vote choice
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = vote, y = age, fill = vote)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Age Distribution by Vote Choice",
       x = "Vote Choice", y = "Age") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")

# ----------------------------------------------------------------------------
# Q8. Correlation matrix
# ----------------------------------------------------------------------------

teds %>%
  dplyr::select(age, edu, income, Taiwanese, Econ_worse, DPP) %>%
  cor(use = "complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "cor") %>%
  ggplot(aes(x = var1, y = var2, fill = cor)) +
  geom_tile() +
  geom_text(aes(label = round(cor, 2)), size = 6) +
  scale_fill_gradient2(low = "#e74c3c", mid = "white", high = "#2980b9",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Matrix of Numeric Variables",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 18) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ----------------------------------------------------------------------------
# Q9. Pairs plot
# ----------------------------------------------------------------------------

teds %>%
  dplyr::select(age, edu, income, Taiwanese, Econ_worse, vote) %>%
  ggpairs(aes(color = vote, alpha = 0.5),
          upper = list(continuous = wrap("cor", size = 4)),
          lower = list(continuous = wrap("points", size = 0.5))) +
  scale_color_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  theme_minimal(base_size = 14)


# ============================================================================
# PART II: UNSUPERVISED LEARNING
# ============================================================================

# ----------------------------------------------------------------------------
# Prepare data for clustering
# ----------------------------------------------------------------------------

teds_numeric <- teds %>%
  dplyr::select(age, edu, income, Taiwanese, Econ_worse, DPP) %>%
  scale() %>%
  as.data.frame()

head(teds_numeric)

# ----------------------------------------------------------------------------
# Q10. Choosing k: Elbow method
# ----------------------------------------------------------------------------

fviz_nbclust(teds_numeric, kmeans, method = "wss", k.max = 10) +
  labs(title = "Elbow Method: Optimal Number of Clusters") +
  theme_minimal(base_size = 18)

# Silhouette method
fviz_nbclust(teds_numeric, kmeans, method = "silhouette", k.max = 10) +
  labs(title = "Silhouette Method: Optimal Number of Clusters") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q11. K-Means clustering
# ----------------------------------------------------------------------------

set.seed(6323)
km_result <- kmeans(teds_numeric, centers = 3, nstart = 25)

table(km_result$cluster)
round(km_result$centers, 2)

# Visualize clusters
fviz_cluster(km_result, data = teds_numeric,
             geom = "point", 
             ellipse.type = "convex",
             palette = c("#e74c3c", "#2980b9", "#27ae60"),
             ggtheme = theme_minimal(base_size = 18)) +
  labs(title = "K-Means Clustering of TEDS 2016 Voters (k = 3)")

# ----------------------------------------------------------------------------
# Q12. Cluster profiling
# ----------------------------------------------------------------------------

teds$cluster <- as.factor(km_result$cluster)

teds %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    mean_age = round(mean(age), 1),
    mean_edu = round(mean(edu), 1),
    mean_income = round(mean(income), 1),
    pct_Taiwanese = round(mean(Taiwanese) * 100, 1),
    pct_Econ_worse = round(mean(Econ_worse) * 100, 1),
    mean_DPP = round(mean(DPP), 2),
    pct_Tsai = round(mean(vote == "Tsai") * 100, 1)
  )

# ----------------------------------------------------------------------------
# Q13. Clusters vs. vote choice
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = cluster, fill = vote)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Vote Choice Distribution by Cluster",
       x = "Cluster", y = "Proportion",
       fill = "Vote") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q14. Principal Component Analysis
# ----------------------------------------------------------------------------

pca_result <- prcomp(teds_numeric, scale. = TRUE)
summary(pca_result)

# Scree plot
fviz_eig(pca_result, addlabels = TRUE) +
  labs(title = "Scree Plot: Variance Explained by Each PC") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q15. PCA variable contributions
# ----------------------------------------------------------------------------

fviz_pca_var(pca_result, 
             col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE) +
  labs(title = "PCA: Variable Contributions") +
  theme_minimal(base_size = 18)

# PCA individuals by vote
fviz_pca_ind(pca_result,
             geom = "point",
             col.ind = teds$vote,
             palette = c("#2980b9", "#27ae60"),
             addEllipses = TRUE,
             legend.title = "Vote") +
  labs(title = "PCA: Voters Projected onto First Two PCs") +
  theme_minimal(base_size = 18)


# ============================================================================
# PART III: SUPERVISED LEARNING
# ============================================================================

# ----------------------------------------------------------------------------
# Q16. Train-test split
# ----------------------------------------------------------------------------

set.seed(6323)

train_index <- createDataPartition(teds$vote, p = 0.7, list = FALSE)
train_data <- teds[train_index, ] %>% dplyr::select(-cluster)
test_data  <- teds[-train_index, ] %>% dplyr::select(-cluster)

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
prop.table(table(train_data$vote))

# ----------------------------------------------------------------------------
# Q17. Logistic regression
# ----------------------------------------------------------------------------

logit_model <- glm(vote ~ age + gender + edu + income + 
                     Taiwanese + Econ_worse + DPP,
                   data = train_data, 
                   family = binomial)

summary(logit_model)

# Evaluation
logit_probs <- predict(logit_model, newdata = test_data, type = "response")
logit_pred <- ifelse(logit_probs > 0.5, "Tsai", "Other")
logit_pred <- factor(logit_pred, levels = c("Other", "Tsai"))

confusionMatrix(logit_pred, test_data$vote)

# ----------------------------------------------------------------------------
# Q18. Decision tree
# ----------------------------------------------------------------------------

tree_model <- rpart(vote ~ age + gender + edu + income + 
                      Taiwanese + Econ_worse + DPP,
                    data = train_data,
                    method = "class")

rpart.plot(tree_model, 
           type = 4, 
           extra = 104,
           main = "Decision Tree: Predicting Vote Choice",
           box.palette = "BuGn",
           cex = 1.2)

# Evaluation
tree_pred <- predict(tree_model, newdata = test_data, type = "class")
confusionMatrix(tree_pred, test_data$vote)

# ----------------------------------------------------------------------------
# Q19. Random forest
# ----------------------------------------------------------------------------

set.seed(6323)

rf_model <- randomForest(vote ~ age + gender + edu + income + 
                           Taiwanese + Econ_worse + DPP,
                         data = train_data,
                         ntree = 500,
                         importance = TRUE)

print(rf_model)

# Evaluation
rf_pred <- predict(rf_model, newdata = test_data)
confusionMatrix(rf_pred, test_data$vote)

# ----------------------------------------------------------------------------
# Q20. Variable importance
# ----------------------------------------------------------------------------

importance_df <- as.data.frame(importance(rf_model)) %>%
  rownames_to_column("variable") %>%
  arrange(desc(MeanDecreaseGini))

ggplot(importance_df, aes(x = reorder(variable, MeanDecreaseGini), 
                           y = MeanDecreaseGini)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest: Variable Importance",
       subtitle = "Mean Decrease in Gini Index",
       x = NULL, y = "Importance") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q21. Model comparison
# ----------------------------------------------------------------------------

results <- tibble(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(
    confusionMatrix(logit_pred, test_data$vote)$overall["Accuracy"],
    confusionMatrix(tree_pred, test_data$vote)$overall["Accuracy"],
    confusionMatrix(rf_pred, test_data$vote)$overall["Accuracy"]
  )
)

results %>% arrange(desc(Accuracy))

# Comparison plot
ggplot(results, aes(x = reorder(Model, Accuracy), y = Accuracy, fill = Model)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(Accuracy * 100, 1), "%")), 
            hjust = -0.1, size = 7) +
  coord_flip() +
  scale_fill_manual(values = c("#e74c3c", "#2980b9", "#27ae60")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(title = "Model Comparison: Prediction Accuracy",
       x = NULL, y = "Accuracy") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")
```

## 3. Use the TEDS2016 dataset to run a multiple regression model

```{r}
library(haven)
library(dplyr)
library(ggplot2)
TEDS_2016 <- read_stata("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")
```

Inspect the data
```{r}
dim(TEDS_2016)
names(TEDS_2016)
summary(TEDS_2016)
```

Keep the variables needed for the assignment

Here I use votetsai as the dependent variable, and age, edu, and income as predictors.

```{r}
teds_ml <- TEDS_2016 %>%
  dplyr::select(votetsai, age, edu, income) %>%
  na.omit()
```

Run a multiple regression model
```{r}
ml_model <- lm(votetsai ~ age + edu + income, data = teds_ml)
summary(ml_model)
```

4. Write a function called regplot to plot a regression line

```{r}
regplot <- function(x, y, xlab = "X", ylab = "Y", main = "Regression Plot") {
  fit <- lm(y ~ x)
  plot(x, y,
       xlab = xlab,
       ylab = ylab,
       main = main,
       pch = 19,
       col = "gray40")
  abline(fit, col = "red", lwd = 2)
}
```

5. Run regplot on the dependent variable using age, education, and income
a. Age
```{r}
regplot(
  x = teds_ml$age,
  y = teds_ml$votetsai,
  xlab = "Age",
  ylab = "Vote for Tsai",
  main = "Vote for Tsai vs. Age"
)
```
b. Education
```{r}
regplot(
  x = teds_ml$edu,
  y = teds_ml$votetsai,
  xlab = "Education",
  ylab = "Vote for Tsai",
  main = "Vote for Tsai vs. Education"
)
```
c. Income
```{r}
regplot(
  x = teds_ml$income,
  y = teds_ml$votetsai,
  xlab = "Income",
  ylab = "Vote for Tsai",
  main = "Vote for Tsai vs. Income"
)
```

6. What is the problem? Why?

The main problem is that the dependent variable is not truly continuous. It is categorical and has only a small number of categories. In this case, votetsai is essentially a binary outcome, indicating whether the respondent voted for Tsai or not. Because linear regression assumes a continuous dependent variable, OLS is not the best model here. The fitted values can fall outside the 0 to 1 range, and the linearity and constant-variance assumptions are not well suited for a binary dependent variable.

This is also consistent with the workshop code, which later recodes the vote variable into a two-category factor and uses logistic regression for prediction rather than ordinary least squares.

7. What can be done to improve prediction of the dependent variable?

A better approach is to use a classification model instead of linear regression. The most natural improvement is logistic regression, because the dependent variable is binary. Logistic regression constrains predicted probabilities to the 0 to 1 range and is more appropriate for modeling vote choice.

For example:
```{r}

logit_model <- glm(votetsai ~ age + edu + income,
                   data = teds_ml,
                   family = binomial)
summary(logit_model)
```
Predicted probabilities
```{r}
teds_ml$pred_prob <- predict(logit_model, type = "response")
head(teds_ml)
```

Simple classification rule
```{r}
teds_ml$pred_class <- ifelse(teds_ml$pred_prob > 0.5, 1, 0)
table(Predicted = teds_ml$pred_class, Actual = teds_ml$votetsai)
```

In addition, prediction may improve by adding more substantively relevant predictors. The workshop code later includes variables such as gender, Taiwanese identity, economic evaluations, and party-related measures, and then compares logistic regression, decision tree, and random forest models. That is a better strategy than relying only on age, education, and income.

Brief conclusion

Using lm() here is helpful as an exercise, but it is not ideal because the dependent variable is categorical. A logistic regression model is more appropriate, and prediction can be further improved by adding relevant political variables and using classification methods.