---
title: "assignment7"
format: html
---

## Assignment 7: AI Exercise

### 1: Shiny App 1 – University Database Instructor Salaries

#### (a) Enter Password to Access PostgreSQL Database

The original `app.R` connects to a local PostgreSQL server.
To access the `university` database, the `password` argument in
`dbConnect()` must be set to your actual PostgreSQL password.

```{r}
#| eval: false

library(shiny)
library(DBI)
library(RPostgres)
library(ggplot2)

ui <- fluidPage(
  titlePanel("EPPS6354 Shiny Workshop 1: University Database - Instructor Salaries"),
  mainPanel(
    plotOutput("salaryPlot")
  )
)

server <- function(input, output, session) {

  # (a) Password entered to connect to local PostgreSQL server
  con <- dbConnect(
    RPostgres::Postgres(),
    dbname   = "university",
    host     = "localhost",
    port     = 5432,
    user     = "postgres",
    password = "yourpassword!"  # Replace with your actual PostgreSQL password
  )

  session$onSessionEnded(function() {
    dbDisconnect(con)
  })

  instructor_data <- reactive({
    query <- "SELECT dept_name, AVG(salary) AS avg_salary
              FROM instructor
              GROUP BY dept_name;"
    dbGetQuery(con, query)
  })

  output$salaryPlot <- renderPlot({
    data <- instructor_data()
    ggplot(data, aes(x = reorder(dept_name, avg_salary), y = avg_salary)) +
      geom_bar(stat = "identity", fill = "steelblue") +
      theme_minimal() +
      labs(
        title = "Average Instructor Salary by Department",
        x = "Department",
        y = "Average Salary"
      ) +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        text = element_text(family = "Palatino")
      )
  })
}

shinyApp(ui = ui, server = server)
```

#### (b) Change Salary Sorting from High to Low

Adding a minus sign inside `reorder()` reverses the sort order to descending (high to low).

```{r}
#| eval: false

library(shiny)
library(DBI)
library(RPostgres)
library(ggplot2)

ui <- fluidPage(
  titlePanel("EPPS6354 Shiny Workshop 1: University Database - Instructor Salaries"),
  mainPanel(
    plotOutput("salaryPlot")
  )
)

server <- function(input, output, session) {

  con <- dbConnect(
    RPostgres::Postgres(),
    dbname   = "university",
    host     = "localhost",
    port     = 5432,
    user     = "postgres",
    password = "yourpassword!"
  )

  session$onSessionEnded(function() {
    dbDisconnect(con)
  })

  instructor_data <- reactive({
    query <- "SELECT dept_name, AVG(salary) AS avg_salary
              FROM instructor
              GROUP BY dept_name;"
    dbGetQuery(con, query)
  })

  # (b) Changed reorder(dept_name, avg_salary) to reorder(dept_name, -avg_salary)
  #     The negative sign sorts bars from high to low
  output$salaryPlot <- renderPlot({
    data <- instructor_data()
    ggplot(data, aes(x = reorder(dept_name, -avg_salary), y = avg_salary)) +
      geom_bar(stat = "identity", fill = "steelblue") +
      theme_minimal() +
      labs(
        title = "Average Instructor Salary by Department (High to Low)",
        x = "Department",
        y = "Average Salary"
      ) +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        text = element_text(family = "Palatino")
      )
  })
}

shinyApp(ui = ui, server = server)
```

#### (c) Try Another Variable from Another Table

```{r}
#| eval: false

library(shiny)
library(DBI)
library(RPostgres)
library(ggplot2)

ui <- fluidPage(
  titlePanel("EPPS6354 Shiny Workshop 1: University Database - Student Credits"),
  mainPanel(
    plotOutput("creditPlot")
  )
)

server <- function(input, output, session) {

  con <- dbConnect(
    RPostgres::Postgres(),
    dbname   = "university",
    host     = "localhost",
    port     = 5432,
    user     = "postgres",
    password = "yourpassword!"
  )

  session$onSessionEnded(function() {
    dbDisconnect(con)
  })

  # (c) Query the student table for avg total credits by department
  student_data <- reactive({
    query <- "SELECT dept_name, AVG(tot_cred) AS avg_cred
              FROM student
              GROUP BY dept_name;"
    dbGetQuery(con, query)
  })

  output$creditPlot <- renderPlot({
    data <- student_data()
    ggplot(data, aes(x = reorder(dept_name, -avg_cred), y = avg_cred)) +
      geom_bar(stat = "identity", fill = "coral") +
      theme_minimal() +
      labs(
        title = "Average Student Total Credits by Department (High to Low)",
        x = "Department",
        y = "Average Total Credits"
      ) +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        text = element_text(family = "Palatino")
      )
  })
}

shinyApp(ui = ui, server = server)
```

### 2: Shiny App 2 – Old Faithful Geyser Data

#### (a) Change the Color of the Bar Chart

```{r}
#| eval: false

library(shiny)

ui <- fluidPage(
  titlePanel("EPPS6354 Shiny Workshop 2: Old Faithful Geyser Data"),
  sidebarLayout(
    sidebarPanel(
      sliderInput(inputId = "bins",
                  label = "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)
    ),
    mainPanel(
      plotOutput(outputId = "distPlot")
    )
  )
)

server <- function(input, output) {

  output$distPlot <- renderPlot({

    x    <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)

    # (a) Changed col from "blue" to "tomato"
    hist(x, breaks = bins, col = "tomato", border = "white",
         xlab = "Waiting time to next eruption (in mins)",
         main = "Histogram of waiting times")

  })
}

shinyApp(ui = ui, server = server)
```
